The four postulates of quantum mechanics are three

### 2020-12-01 Articles/2019-QuantumComposition/Submissions/PhysLetters/response.tex

\documentclass[11pt]{article}

\usepackage[margin=1.25in]{geometry}

\usepackage{amssymb}
\usepackage{color}
\usepackage{graphicx}
\usepackage{epsfig,amssymb,amsmath,amsthm}
%\usepackage[active]{srcltx}
%\usepackage[hypertex,linkcolor=red]{hyperref}

\usepackage{color}
\usepackage{cancel}
\usepackage{tikz-cd}

\DeclareMathOperator{\spn}{span}

\newcommand{\pj}[1] {\underbar{$#1$}}
%\newcommand{\pj}[1] {\overline{#1}}

\def\>{\rangle}
\def\<{\langle}
\def\ca{_{\cal A}}
\def\cb{_{\cal B}}
\def\cc{_{\cal C}}
\def\comment#1{}
%\def\comment#1{ [{\bf Comment Lor:} {\sf #1}]}
\def\commentg#1{ [{\bf Comment Gabriele:} {\sf #1}]}
\def\labell#1{\label{#1}}
%\def\labell#1{\label{#1}{\mbox{{\tiny #1}}}}
%\def\section#1{{\par\em #1:--- }}
\def\togli#1{}
\def\sh{\mbox{sh}}
\def\iden{\openone}

\begin{document}

Dear Dr. Robert Garisto,

thank you very much for having provided us
good quality referee reports for our paper. We appreciate that this is
not an easy task.

We believe we have overcome the criticisms pointed out by the referees
and we detail below our point-by-point response. Referee A suggested
that our paper ``could be published in PRL after substantial re-writing
and polishing of the arguments", which we have carefully done
following his/her useful suggestions, even though it took a
substantial effort to rewrite most of the paper. Referee B appreciates
our work (``There is no doubt that this paper represents some fine
mathematical work"), but questions its suitability for PRL. However,
the Referee has essentially focused on the supplementary material, in
his/her words: ``The essence of the paper, to my mind, lies in the
supplementary information." We agree with the Referee that the
supplementary information is not suitable as a paper for PRL. However,
the paper itself was carefully written to emphasize the PHYSICAL
arguments whereas the supplementary material just contains the
(nontrivial) mathematical details behind the physics. We urge the
Referee to take a fresh look at the paper with this in mind. Of course
we take full responsibility for the Referee's misconception: the
Referee is clearly very qualified, and in our previous version of the
paper we had failed to make this distinction clear. (Indeed we had
wrongly claimed that the supplementary material contained the same
material formulated in a different manner.)

Thank you for your time.

In the following we include our detailed answers to the Referees. Note 
that we will not enclose the changes made to
the paper, as we have modified most of it as requested by Referee A (removing some parts to make
space).

\section*{Referee A}

\begin{quote}
The authors address a very interesting problem that has
not received much attention, and they present an important result
about the internal logical structure of quantum theory.
\end{quote}

We thank the Referee for the positive assessment of our work.
\begin{quote}
I would say that the result is correct, but I could not
completely follow the argumentation of the article. I found the text
quite confusing and I think that it does not meet the writing
standards of PRL. In my opinion the arguments could be simplified and
expressed with simpler mathematics (see below). I think that the paper
could be published in PRL after substantial re-writing and polishing
of the arguments.
\end{quote}

We appreciate the suggestion that our paper can be published in
PRL. As suggested, we have substantially rewritten our paper following
the useful and insightful suggestions by the Referee, which we
sincerely thank for the feedback. We should remark that this was not
an easy undertaking, and we had to devote substantial effort in
rewording most of the paper. Nonetheless, we believe it was worth it,
because the argument is much clearer now.

\begin{quote}
1. Regarding 'statistical independence is not an
additional requirement: it is already contained in the measurement
postulate.' I think that this is a bit more subtle. [...]
\end{quote}

This is a great argument! We had not thought of that. The Referee is
correct, one can indeed prove the statistical independence of the
outcomes as the Referee does. And indeed the Referee is correct that
we had not considered that.

However, upon re-checking our proof very carefully, we realized that
we do not need the statistical independence of the outcomes. We just
need the statistical independence of the preparations. Nonetheless, we
believe that the argument the Referee proposes is really nice and we
respectfully ask the Referee if we can add it to our paper (citing an
anonymous Referee, or the Referee may contact us separately if he/she
wants to be properly acknowledged). For now we have added this at the end of the
supplementary material, but, if the Referee disagrees, we can remove
it and replace it with a sentence of the type: ``The statistical
independence of the outcomes is not an additional assumption, as it
can be proved independently."

The Referee's comment made us realize that we didn't discuss the
statistical independence in sufficient detail, and that is actually
one of the critical step in the demonstration (i.e. making explicit an
implicit assumption). We rewrote and expanded the part of the paper
(and the supplementary material) where we discuss this, mainly in the
first column of page 3.

\begin{quote}
2. I would say that physicists are familiar with the
notion of 'colinear'.  Hence, I would either (i) provide a definition
or (ii) express the argument with the more familiar notions 'linear'
and 'anti-linear' I would be inclined for the second (if possible).
\end{quote}

We presume that the Referee intended ``physicists are UNfamiliar".  We
agree and we also agree that we have not done enough in the text to
guide the reader in that respect. As suggested, we have explained what
``colinear" means. As far as we understand, we cannot remove this
notion from our proof, unfortunately. Indeed, the use of projective
geometry is necessary. We now have a paragraph that clearly says that
and why we need projective geometry.

This is necessary to understand the properties of the maps M and m
that we need to derive the tensor products.

\begin{quote}
3. Regarding 'the dual-space description of a quantum
system through bras is equivalent to the description using kets', what
means 'equivalent' here? Can you justify it with an argument instead
of the anecdote that 'in his first papers Schroedinger used both signs
in his equation, effectively writing two equivalent equations with
complex-conjugate solutions'?
\end{quote}

We simply meant the following. Given any equation in quantum mechanics
in terms of operators acting on ket vectors, one can write an
EQUIVALENT equation in terms of bra vectors by simply taking the
Hermitian conjugate of both sides of the first equation. Namely, the
equation $A|\psi\>=B|\phi\>$ is equivalent to $\<\psi|A^\dag=\<\phi|B^\dag$. In
terms of wave functions, it means taking the complex conjugate (which
entails a change of sign in the case of the Schroedinger equation). We
apologize this was unclear (upon re-reading that part, we agree) and
we have now specified it explicitly. We rewrote that part completely,
and we added the previous equivalence explicitly.

\begin{quote}
4. Regarding 'The function m, then, is an isomorphism over
all elements of the basis $|a_i\>|b_j\>$'. What it means to be an
'isomorphism over' some vectors?
\end{quote}

We mean that the function m is a one-to-one mapping, that maps the
elements of the basis into elements of a basis of a tensor product
space. We understand our notation could lead to confusion. We have
rewrote that part, and also changed the (admittedly confusing)
notation to clarify the whole thing.

\begin{quote}
5. Regarding 'other linear products of linear spaces
exist'. Is the direct product of two vector spaces a vector space? (I
am assuming that a composite system is also a system and hence it has
an associated Hilbert space.)
\end{quote}

Yes, a direct product of vector spaces is a vector space. You can
certainly define a Hilbert space using direct products of vectors of
two Hilbert spaces (you just need to make sure the norm doesn't
diverge). Of course, that is not the correct thing to do in quantum
mechanics. We are simply pointing out that there are many ways to map
two linear spaces to another linear space, so requiring linearity on
the spaces is not sufficient to single out the tensor product. In
fact, linearity by itself is not sufficient to warrant use of a
product to create the joint space.

We have changed that sentence to make it clearer, and have also added
a footnote to further clarify the distinction

\begin{quote}
How you define multiplication by scalars? If it is defined
distributively then this direct product becomes the direct sum of the
two spaces. And similarly with the topological product. Therefore,
these three examples of product of linear spaces are the same one.
\end{quote}

Each linear space comes with its own sum (e.g. $+ : A \times A \to A$)
and its own scalar multiplication (e.g. $* : \mathbb{R} \times A \to
A$). Distributivity is $r * (a + b) = (r*a) + (r*b)$. Each space
individually will retain that property.

If we imagine a map $m : A \times B \to C$, we can ask how can we
express $r*m(a,b)$ as a function of $m(r*a)$ and $m(r*b)$. This
differs from product to product. For the direct product, we have $r *
(a \times b) = r*a \times r*b$ which is distributive. For the tensor
product, we have $r * (a \otimes b) = r*a \otimes b = a \otimes r*b$
which makes it not distributive.

We can also ask how we can express $m(a,b+c)$ in terms of $m(a,b)$ and
$m(a,c)$. For the tensor product we have $a \otimes (b+c) = a \otimes
b + a \otimes c$ which is distributive. For the direct product we have
$a \times (b+c) = a \times b + 0 \times c$ where $0$ is the zero
vector. So the direct product is not distributive.

The direct sum and direct product are the same in the finite case but
not in the infinite case. In that case, one can only take finitely
many non-zero vectors for the direct sum, while this is not required
in the direct product.

The overall point is: there are many choices. For distributivity, one
has to decide if it's distributivity over the addition or the scalar
multiplication. That choice needs to be motivated. The exterior
product is also distributive over the addition, like the tensor
product (as it is a quotient of the tensor product). So, why the
tensor product and not the exterior product?

We added the following footnote, hoping it will succinctly highlight
the issue: For example, in the tensor product $a \otimes (b+c) = a
\otimes b + a \otimes c$ while in the direct product $a \times (b+c) =
a \times b + 0 \times c$ where $0$ is the zero vector. Also, in the
tensor product $r * (a \otimes b) = (r*a) \otimes b = a \otimes (r*b)$
while in the direct product $r * (a \times b) = r*a \times r*b$.

This also addresses the next point.

\begin{quote}
6. Regarding '(Dirac) adds the seemingly innocuous
request'. I understand that requiring the product of vectors to be
distributive is the same as requiring it to be bilinear, that is $|a\>
(|b\> + |c\>) = |a\>|b\> + |a\>|c\>$. Am I missing something?
\end{quote}

No, what the Referee states is correct. Our point is that such request
is far from innocuous, as for example it excludes (without any
physical justification whatsoever) the direct product. The requirement
of distributivity (or bilinearity) introduced by Dirac is an
ADDITIONAL assumption that is certainly not warranted and it is even
NOT natural. For example, it does not hold in classical mechanics,
where the state space of composite systems is obtained through the
direct product, and (a,b+c) is DIFFERENT from (a,b)+(a,c).

\begin{quote}
Supplementary information 1. Definition of composite
system. Why do you say `for any compatible pair' instead of `for any
pair'? Can we have a pair of systems where not all pair of states are
compatible?
\end{quote}

We can have a pair of systems where not all pairs of states are
compatible. Consider two electrons. They cannot be prepared in the
same state. Therefore the state space of the composite system should
exclude those cases. We use the exterior product (i.e. the
anti-symmetrized tensor product) in those cases. As we explicitly say
elsewhere, electrons cannot be properly be considered as quantum
subsystems (at least in the case in which indistinguishability plays a
role): the quantum system is the field, and electrons are just field
excitations (quanta).

We now made this more explicit after the definition: Requirement 1
ensures that the composite system is well defined at least when the
components are prepared independently. Conceptually, this ensures that
the composite system contains all the properties of the
components. Note that superselection rules or other restrictions may
prevent the independent preparation of all possible pairs (e.g.~two
fermions cannot be jointly prepared in the same state). The tensor
product is recovered only when all pairs are compatible.

\begin{quote}
``I would say that this high degree of generality obscures
the argument. Perhaps this case of systems with non-compatible pairs
of states can be left for a discussion after the main proof."
\end{quote}
We agree with the Referee, but we suppose that the Supplementary
information section will be only read by people who want to understand
the full details of our argument. This type of readers will not be
scared by the full generality of the argument. Instead in the main
paper we were more careful to consider first the compatible states and
then the general issues. In the Supplementary information we need to
consider the situation in full generality, e.g.  systems of multiple
fermions and bosons when considered as systems as is sometimes done
improperly (a common enough case that we have to justify why, in those
cases, the tensor product is not used directly). In the proof, we
already concentrate on the preparation independent case as we also
didn't want to make the derivation too heavy. We hope that the above
clarification explains why we believe we have to make that distinction
mathematically and why it is also interesting physically. In any case,
we have clarified this point both in the paper and the supplementary
material.

\begin{quote}
2. Regarding 'the proposition a wedge b is equivalent to
the (pure) state $M (a, b)$'. The authors do not define what it means
for a logical proposition to be equivalent to a ray? Is it necessary
to use the formalism of logical propositions to derive the tensor
product? If so, some definitions should be introduced.  Similarly,
what it means to say that a state $M (a, b)$ corresponds to an event in
probability space. I have the feeling that the same can be said in
more plain words.
\end{quote}

We agree that we haven't clearly explained the formalism and we have
now done so. We do not really need the formalism of logical
propositions, but just the logical connective ``AND" that we indicated
with the wedge symbol. That's all we need, and it's necessary for
defining the JOINT probability. We could remove the wedge symbol and
replace it with the connective "and". We tried, but the formulas
become quite heavy, so we believe that this (minimal) concession to
the formalism is useful.

We have modified the paper (and the supplementary material) in
multiple places to clarify the formalism: not only the formalism of
logical propositions, but also how these are mapped in the formalism of quantum mechanics
(e.g. rays and Hilbert subspaces).

\begin{quote}
3. Regarding `Every preparation of the composite gives a
nontrivial measurement on the components.' Do you mean that, for any
state of the composite system, local projective measurements must have
at least one outcome with non-zero probability? I think that this
phrasing is easier to accept.
\end{quote}
Yes. That is exactly what we meant. We agree that the Referee's
wording is clearer and we have replaced it in the new version of our
paper.

In conclusion, we thank the Referee for the careful reading and the
thoughtful suggestions. They have been all very valuable in enhancing
the paper.

\section*{Referee B}

\begin{quote}
This paper shows how the tensor-product assumption of
quantum theory can be derived from two standard postulates of quantum
theory--specifically, the postulates about states and
measurements--together with a set of assumptions reflecting reasonable
interpretations of the expressions 'composite system' and 'independent
systems'. The essence of the paper, to my mind, lies in the
supplementary information.
\end{quote}

We respectfully disagree with this assessment: while the supplementary
information contains the nontrivial mathematical formulation of the
ideas presented in our paper, we believe that our main contribution is
in the PHYSICS, rather than in the MATHEMATICS. Namely, we show how
some PHYSICAL ASSUMPTIONS (contained in the other postulates of
quantum mechanics) must give rise to tensor products. We have
carefully written our paper on two separate levels (actually three):
at a first level, a physicist interested in the main physical ideas
will only read the paper, and can easily skip the sketch of the proof
presented there, while still understanding the main results. At a
second level, a reader that wants more details can also follow the
sketch of the proof that we put in the main paper. Finally, a
specialist of quantum mechanics (such as, clearly, the Referee) can
get into the fine details of the argument by looking at the physical
considerations in the paper and the mathematical details in the
supplementary information. We understand that this distinction was not
clear in the first version of the paper, and take we take
responsibility for this (we ourselves had mischaracterized our
supplementary information). We urge the Referee to re-read the paper
keeping in mind that the main paper is written for a more general
audience and that mostly contains the physics, and the supplementary
information only contains the details of the mathematical proofs.
\begin{quote}
This is where one finds a precise statement of each
assumption and a precise articulation of the logical connections
between statements. Indeed, the authors' main contribution is the
articulation of this logical argument. A number of physicists may
already hold the intuition that the tensor product rule is the only
composition rule one can imagine in light of the rest of the theory;
however, making this intuition precise, as the authors have done, is
another matter.\end{quote}

We thank the Referee for this compliment. We agree: most physicists do
hold the intuition that the tensor product is the natural composition
rule for quantum mechanics, but, as the Referee suggests, this had
indeed never been shown previously, at least using the conventional
formulation of quantum mechanics as we do.

\begin{quote}There is no doubt that this paper represents some fine
mathematical work.\end{quote}

Thank you!

\begin{quote}The question is whether it meets the criteria for
publication in PRL.  As the authors point out, the tensor product
postulate has been addressed by other authors. Some authors explain
why this postulate is reasonable but do not prove in detail how it
follows from other assumptions. Within the quantum logic community,
the tensor product rule has been derived rigorously from simpler
axioms. The authors' approach is different in that they are starting
with two other standard postulates of quantum theory and thereby aim
to show that there is some redundancy in the standard postulates."
\end{quote}
We agree: this is a good characterization of our work.

\begin{quote}In this respect, the authors' approach could be compared
to that of Ref. [8], which derives the measurement postulate from
other standard postulates (including the tensor product
postulate). Ref. [8] derives not only the mathematical representation
of measurement outcomes, but also the Born rule and the general form
of the rule for determining the post-measurement state of the
system. That paper was published in Nature Communications. Another
paper, Ref. [28], which also derives the Born rule from the postulate
about states and the tensor product rule, was published many years ago
in American Journal of Physics.
\end{quote}

It is true that both our result and theirs shows that there is a certain redundancy in the postulates, even though our result is completely different from theirs. So, following the Referee's suggestion we have emphasized this.

\begin{quote}However, I think there is an important difference between
the tensor product postulate and the measurement postulate. The
measurement postulate is the central focus of the notorious
measurement problem of quantum theory, which has been addressed in
many books and papers and which many would argue is the root of all
the interpretational difficulties. (Indeed, Ref. [28] explicitly
addresses the measurement problem.) The tensor product postulate does
not have this status. So, whereas a paper deriving the measurement
postulate might well be read by any physicist with even a passing
interest in the conceptual foundations of the theory, a paper deriving
the tensor product postulate will not have the same breadth of
appeal. Also, I think it is somehow less surprising that the tensor
product postulate emerges in this way than that the measurement
postulate does. The most obvious alternative to the tensor product is
the direct sum, which one immediately sees does not capture what is
meant by 'composite system'.\end{quote}

We certainly agree with the Referee that a solution of the measurement
problem would be more interesting than our result. BUT it is quite
debatable (and actually debated) whether any of the above results
resolved the measurement problem: the general consensus is that it is
still OPEN. In contrast, our result is quite solid (and neither
Referee objects to that): we DO show that the tensor product postulate
is not a necessary postulate, and believe (more on this below) that
our result is certainly suitable for publication on PRL. Referee A
shares this belief.

### 2020-12-01 Articles/2019-QuantumComposition/Submissions/PhysLetters/tensor.tex
After Lorenzo's review

	mechanics: the state postulate and the measurement postulate.
	The tensor product postulate does not appear in all axiomatizations of

@@

	products of linear spaces exist, such as the direct product, the exterior/wedge product or the direct sum of vector spaces, which is used
	in classical mechanics to combine state spaces of linear systems. These are all maps from linear spaces to linear spaces but they differ in how the linearity of one is mapped to the linearity of the other.\footnote{For example, in the tensor product $a \otimes (b+c) = a \otimes b + a \otimes c$ while in the direct product $a \times (b+c) = a \times b + 0 \times c$ where $0$ is the zero vector. Also, in the tensor product $r  (a \otimes b) = (r a) \otimes b = a \otimes (r b)$ while in the direct product $r  (a \times b) = r a \times r b$, where $r$ is a scalar.}
	This belief may have arisen from the seminal book of Dirac
	\cite{diracbook}, who introduces tensor products (Chap.~20) by

@@

	linear functions of them). It is not an innocuous request. For
	example it does not hold where the composite vector space of two
	linear spaces is described by the direct product, e.g.~in classical
	mechanics, for two strings of a guitar: it is not distributive.
	[General classical systems, not only linear ones, are also composed
	through the direct product.] Of course, Dirac is not constructing an
	axiomatic formulation, so his `sleight of hand' can be forgiven. In
	contrast, von Neumann (\cite{vonneumannbook} Chap.~VI.2, also
	\cite{jauch}) introduces tensor products by noticing that this is a
	natural choice in the position representation of wave mechanics (where
	they were introduced in \cite{weyl,epr}), and then {\em explicitly
		postulates} them in general: ``This rule of transformation is
	correct in any case for the coordinate and momentum operators [...]
	and it conforms with the [observable axiom and its linearity
	principles], we therefore postulate them generally.''
	\cite{vonneumannbook}.  More mathematical or conceptually-oriented
	modern formulations (e.g.~\cite{ozawa,masanes,wootters,nielsenchuang})
	introduce this postulate explicitly.  An interesting alternative is
	provided in \cite{ballentinebook,ballentinepaper}: after introducing
	tensor products, Ballentine verifies a posteriori that they give the
	correct laws of composition of probabilities. Similarly, Peres uses

@@

	other postulates. Similar approaches appear in \cite{marmo,aerts}
	where tensor products were obtained by specifying additional physical
	or mathematical requirements. \comment{ {\em Il resto della frase da spostare in supp material} In quantum field theory one tends to
		avoid problems connected with tensor products of infinite dimensional
		spaces by focusing on algebraic commutation structures,
		e.g.~\cite{giddins,roos}.  In particular, the recent MIP*=RE result
		\cite{mipre} implies that, in infinite dimensions, the tensor product
		is strictly less computationally powerful than the commutation
		structures, emphasizing the difference among these two structures, at
		least for the infinite-dimensional case. We will consider the
		non-relativistic setting here.}

	We start
	from the natural definition of a composite system as the set of two
	(or more) quantum systems. The composite system is therefore
	made of system $A$ {\em and} (joined with) system $B$ and {\em nothing else}.\footnote{Note that this does not exclude entangled states since they too are characterized by the observables and only the observables of the constituents.} The first key insight is that the first two postulates of quantum theory (introduced below) already assume that the preparation of one system is independent from the preparation of another (statistical independence). In fact, we cannot even talk about a system in the first place if we cannot characterize it independently. The second key insight is that, since all measurement
	outcomes and all states of the subsystems plus the composite live in the same probability space, we will automatically have a map $M$ that for a pair of subsystems' states gives us the composite state that corresponds to the statistically independent case. These insights are enough to characterize mathematically the state space of the composite: the linearity given by the Hilbert space, together with the fact that the composite system is fully described by the observables of $A$ and $B$, allows to extend the construction to the composite states that do not satisfy statistical independence (the entangled states). So the work consists of two interrelated efforts: a physical argument that starts from the first two postulates and leads to the necessary existence of the composition map $M$ and its properties together with a formal argument that shows how $M$ leads to the tensor product.

@@

	To make these arguments precise, we will need to use elements of projective geometry.
	If $\psi$ is a vector in our Hilbert space, the pure states is represented by the ray $\pj{\psi}$, which is the subspace spanned by $\psi$. This includes not only vectors with arbitrary phase, but also arbitrary modulus. The term ``ray'' can therefore be misleading since $\pj{\psi}$ is really a complex { \em plane }. In the same way, constraining the observable $X$ to a particular value $x_0$ means identifying the subspace spanned by all eigenstates such that $X | \psi \> = x_0 |\psi\>$. As we need to map states and observables between subsystems and composite system, the map $M$ derived from the probability space is really a map between subspaces, between the projective spaces. The technical work, then, is to show that the map $M$ on the projective spaces corresponds to a map $m$ on the vector spaces and that $m$ can be taken to be, without loss of generality, the tensor product.

	This is done by showing that preparation independence and
	statistical independence imply three conditions on the map $m$:
	(H1)~totality: the map is defined on all states of the subsystems;
	(H2)~bilinearity: the map is bilinear thanks to the fundamental
	theorem of projective geometry; (H3)~span surjectivity: the span of
	the image of map coincides with the full composite Hilbert space.  We
	then prove that, if these three conditions H1, H2 and H3 hold, then
	the map $m$ is the tensor product, namely the Hilbert space of the
	composite system is a tensor product of the components: the tensor
	product ``postulate'', which hence loses its status of a postulate. An
	overview of all these logical implications is given in
	Fig.~\ref{f:fig}. The rest of the paper contains the sketch of this
	argument. The supplementary material contains the mathematical details.

@@

	We start with the axiomatization of quantum mechanics based on the following
	postulates (e.g.~\cite{ozawa,masanes,wootters,nielsenchuang}): (a)~The pure state of a
	system is described by a ray $\pj{\psi}$ corresponding to a set of
	non-zero vectors $|\psi\>$ in a complex Hilbert space, and the
	system's observable properties are described by self-adjoint operators
	acting on that space; (b)~The probability that a measurement of a
	property $X$, described by the operator with spectral decomposition \comment{Normalizzare qui e in altri posti!}
	$\sum_{x,i}x|x_i\>\<x_i|$ ($i$ a degeneracy index), returns a value
	$x$ given that the system is in state $\pj{\psi}$ is
	$p(x|\psi)=\sum_i|\<\psi|x_i\>|^2$ (Born rule). (c)~The state
	space of a composite system is given by the tensor product of the
	spaces of the component systems; (d)~The time evolution of an isolated
	system is described by a unitary operator acting on a vector
	representing the system state, $|\psi({t})\>=U_{t}|\psi({t}=0)\>$ or,
	equivalently, by the Schr\"odinger equation. The rest of quantum
	theory can be derived from these axioms. While some axiomatizations
	introduce further postulates, we will be using only (a) and (b) to
	derive (c), so the above are sufficient for our aims.

@@

	The definition of a composite system as containing {\em only} the
	collection of the subsystems means that any preparation of both
	subsystems independently must correspond to the preparation of the
	composite system. Since states are defined by postulate (a) as rays in
	the respective Hilbert spaces, there must exist a map $M : \pj{\cal A}
	\times \pj{\cal B } \to \pj{\cal C}$ that takes a pair of states for
	the subsystems ($\pj{\cal A}$ and $\pj{\cal B }$ represent the
	projective spaces and the Cartesian product is the set of all possible
	pairs) and returns a state in the projective space $\pj{\cal C}$ for
	the composite. To visualize the geometrical meaning of $M$ directly within the Hilbert spaces, given a ray (a complex plane) in each of $\cal A$ and $\cal B$, $M$ returns a ray (a complex plane) in $\cal C$.  Our final goal will be to find a map
	$m:{\cal A}\times{\cal B}\to{\cal C}$ that acts on vectors in the
	Hilbert spaces $\cal A$, $\cal B$ and $\cal C$ consistently with $M$. Namely,
	$\pj{m(a,b)}=M(\pj{a},\pj{b})$ where the underline sign indicates the
	elements in the projective space. Again geometrically, $m$ takes a vector in each of $\cal A$ and $\cal B$, and returns a vector in $\cal C$ and we want this to be consistent with $M$ such that vectors picked from the same rays will return vectors in the same ray. We will prove that the map $m$ is
	the tensor product. We focus on pure states here: the argument can be
	extended to mixed states using standard tools \cite{ballentine}.

@@

	\vskip 1\baselineskip A few comments on the proof: it is based on the
	universal property of the tensor product, which uniquely characterizes
	it. First we show that the bilinear map $m$ maps subsystems' bases
	into the composite system basis. We also know that there exists a
	tensor product map $ {T}=\otimes$ that can compose the vectors in
	$\cal A$ and $\cal B$.  Then we use the universal property: since $m$
	is a bilinear map, we are assured that there exists a unique $\hat m$
	such that $\hat m \circ T=m$. Since we show that $\hat m$ is an
	isomorphism, then $\hat{m}$ bijectively maps vectors in $\cal C$ onto
	vectors in the tensor product space. Namely $m={T}=\otimes$.

>>>>

	mechanics: the state postulate and the measurement postulate.  The
	tensor product postulate does not appear in all axiomatizations of

@@

	products of linear spaces exist, such as the direct product, the
	exterior/wedge product or the direct sum of vector spaces, which is
	used in classical mechanics to combine state spaces of linear systems.
	These are all maps from linear spaces to linear spaces but they differ
	in how the linearity of one is mapped to the linearity of the others
	\footnote{For example, in the tensor product
		$a \otimes (b+c) = a \otimes b + a \otimes c$ while in the direct
		product $a \times (b+c) = a \times b + 0 \times c$ where $0$ is the
		zero vector. Also, in the tensor product
		$r (a \otimes b) = (r a) \otimes b = a \otimes (r b)$ while in the
		direct product $r (a \times b) = r a \times r b$, where $r$ is a
		scalar.}.  This belief may have arisen from the seminal book of
	Dirac \cite{diracbook}, who introduces tensor products (Chap.~20) by

@@

	linear functions of them). It is not an innocuous request. For example
	it does not hold where the composite vector space of two linear spaces
	is described by the direct product, e.g.~in classical mechanics, for
	two strings of a guitar: it is not distributive.  [General classical
	systems, not only linear ones, are also composed through the direct
	product.] Of course, Dirac is not constructing an axiomatic
	formulation, so his `sleight of hand' can be forgiven. In contrast,
	von Neumann (\cite{vonneumannbook} Chap.~VI.2, also \cite{jauch})
	introduces tensor products by noticing that this is a natural choice
	in the position representation of wave mechanics (where they were
	introduced in \cite{weyl,epr}), and then {\em explicitly postulates}
	them in general: ``This rule of transformation is correct in any case
	for the coordinate and momentum operators [...]  and it conforms with
	the [observable axiom and its linearity principles], we therefore
	postulate them generally.''  \cite{vonneumannbook}.  More mathematical
	or conceptually-oriented modern formulations
	(e.g.~\cite{ozawa,masanes,wootters,nielsenchuang}) introduce this
	postulate explicitly.  An interesting alternative is provided in
	\cite{ballentinebook,ballentinepaper}: after introducing tensor
	products, Ballentine verifies a posteriori that they give the correct
	laws of composition of probabilities. Similarly, Peres uses

@@

	other postulates.  In \cite{marmo,aerts} tensor products were obtained
	by specifying additional physical or mathematical requirements.
	\comment{ {\em Il resto della frase da spostare in supp material} In
		quantum field theory one tends to avoid problems connected with
		tensor products of infinite dimensional spaces by focusing on
		algebraic commutation structures, e.g.~\cite{giddins,roos}.  In
		particular, the recent MIP*=RE result \cite{mipre} implies that, in
		infinite dimensions, the tensor product is strictly less
		computationally powerful than the commutation structures,
		emphasizing the difference among these two structures, at least for
		the infinite-dimensional case. We will consider the non-relativistic
		setting here.}

	We start from the natural definition of a composite system as the set
	of two (or more) quantum systems. The composite system is therefore
	made of system $A$ {\em and} (joined with) system $B$ and {\em nothing
		else}. The first key insight is that the first two postulates of
	quantum theory (introduced below) already assume that the preparation
	of one system is independent from the preparation of another
	(statistical independence). In fact, we cannot even talk about a
	system in the first place if we cannot characterize it independently.
	The second key insight is that, since all measurement outcomes and all
	states of the subsystems plus the composite live in the same
	probability space, we will automatically have a map $M$ that for a
	pair of subsystems' states gives us the composite state that
	corresponds to the statistically independent case. These insights are
	enough to characterize mathematically the state space of the
	composite: the linearity given by the Hilbert space, together with the
	fact that the composite system is fully described by the observables
	of $A$ and $B$, allows to extend the construction to the composite
	states that do not satisfy statistical independence (the entangled
	states). So the work consists of two interrelated efforts: a physical
	argument that starts from the first two postulates and leads to the
	necessary existence of the composition map $M$ and its properties
	together with a formal argument that shows how $M$ leads to the tensor
	product.

@@

	This map $M$ acts on the state spaces of the subsystems. Each pure
	state is identified by a ray $\pj{\psi}$, a subspace of the system's
	Hilbert space spanned by all vectors $\psi$ differing by their
	(nonzero) modulus and phase: a two-dimensional subspace (a plane). In
	the same way, constraining the observable $X$ to a particular outcome
	value $x_0$ means identifying the subspace spanned by all
	non-normalized eigenvectors $|x_0\>$ of arbitrary phase such that
	$X | x_0 \> = x_0 |x_0\>$. As we need to map states and observables
	between subsystems and composite system, the map $M$ derived from the
	probability space is really a map between subspaces, namely the
	projective spaces. We then use projective geometry to show that the
	map $M$ on the projective spaces corresponds to a map $m$ on the
	vector spaces which is the tensor product.

	This is done by showing that preparation independence and statistical
	independence of the systems imply three conditions on the map $m$:
	(H1)~totality: the map is defined on all states of the subsystems;
	(H2)~bilinearity: the map is bilinear thanks to the fundamental
	theorem of projective geometry; (H3)~span surjectivity: the span of
	the image of map coincides with the full composite Hilbert space.  We
	then prove that, if these three conditions H1, H2 and H3 hold, then
	the map $m$ is the tensor product, namely the Hilbert space of the
	composite system is a tensor product of the components: the tensor
	product ``postulate'', which hence loses its status of a postulate. An
	overview of all these logical implications is given in
	Fig.~\ref{f:fig}. The rest of the paper contains the sketch of this
	argument, including all the physical arguments outlined above. The
	supplementary material contains the mathematical details.

@@

	We start with the axiomatization of quantum mechanics based on the
	following postulates
	(e.g.~\cite{ozawa,masanes,wootters,nielsenchuang}): (a)~The pure state
	of a system is described by a ray $\pj{\psi}$ corresponding to a set
	of non-zero vectors $|\psi\>$ in a complex Hilbert space, and the
	system's observable properties are described by self-adjoint operators
	acting on that space; (b)~The probability that a measurement of a
	property $X$, described by the operator with spectral decomposition
	$\sum_{x,i}x|x_i\>\<x_i|/|\<x_i|x_i\>|$ ($i$ a degeneracy index),
	returns a value $x$ given that the system is in state $\pj{\psi}$ is
	$p(x|\psi)=\sum_i|\<\psi|x_i\>|^2/|\<x_i|x_i\>\<\psi|\psi\>|$ (Born
	rule). (c)~The state space of a composite system is given by the
	tensor product of the spaces of the component systems; (d)~The time
	evolution of an isolated system is described by a unitary operator
	acting on a vector representing the system state,
	$|\psi({t})\>=U_{t}|\psi({t}=0)\>$ or, equivalently, by the
	Schr\"odinger equation. The rest of quantum theory can be derived from
	these axioms. While some axiomatizations introduce further postulates,
	we will be using only (a) and (b) to derive (c), so the above are
	sufficient to our aims.

@@

	The definition of a composite system as containing {\em only} the
	collection of the subsystems means that any preparation of both
	subsystems independently must correspond to the preparation of the
	composite system. Since states are defined by postulate (a) as rays in
	the respective Hilbert spaces, there must exist a map
	$M : \pj{\cal A} \times \pj{\cal B } \to \pj{\cal C}$ that takes a
	pair of states for the subsystems ($\pj{\cal A}$ and $\pj{\cal B }$
	represent the projective spaces and the Cartesian product is the set
	of all possible pairs) and returns a state in the projective space
	$\pj{\cal C}$ for the composite. To visualize the geometrical meaning
	of $M$ directly within the Hilbert spaces, given a ray (a complex
	plane) in each of $\cal A$ and $\cal B$, $M$ returns a ray (a complex
	plane) in $\cal C$.  Our final goal will be to find a map
	$m:{\cal A}\times{\cal B}\to{\cal C}$ that acts on vectors in the
	Hilbert spaces $\cal A$, $\cal B$ and $\cal C$ consistently with $M$.
	Namely, $\pj{m(a,b)}=M(\pj{a},\pj{b})$ where the underline sign
	indicates the elements in the projective space. Again geometrically,
	$m$ takes a vector in each of $\cal A$ and $\cal B$, and returns a
	vector in $\cal C$ and we want this to be consistent with $M$ such
	that vectors picked from the same rays will return vectors in the same
	ray. We will prove that the map $m$ is the tensor product. We focus on
	pure states here: the argument can be extended to mixed states using
	standard tools \cite{ballentinebook}.

@@

	\vskip 1\baselineskip A few comments on the proof: it is based on the
	universal property of the tensor product, which uniquely characterizes
	it. In step 1 we show that the bilinear map $m$ maps subsystems' bases
	into the composite system basis. We also know that there exists a
	tensor product map $ {T}=\otimes$ that can compose the vectors in
	$\cal A$ and $\cal B$.  In step 2 we use the universal property: since
	$m$ is a bilinear map, we are assured that there exists a unique
	$\hat i$ such that $\hat i \circ T=m$. Since we show that $\hat i$ is
	an isomorphism, then $\hat{i}$ bijectively maps vectors in $\cal C$
	onto vectors in the tensor product space. Namely $m={T}=\otimes$.
